---
title: "Importing, converting and plotting Dates and Times with R"
author: "Betsy Cowdery"
date: "January 25, 2016"
output:
html_document:
code_folding: hide
highlight: tango
---
<STYLE TYPE="text/css">
<!--
tr {
border: 1px solid #666;
}
th {
font-weight: bold;
text-align: Left;
background-color: #cce5ff;
padding: 2px 4px;
border: 1px solid #666;
white-space: nowrap;
}
td{
text-align: Left;
padding: 2px 4px;
border: 1px solid #666;
}
table { 
margin: 1em 0;
border-collapse: collapse;
}
h1:after{
content:' ';
display:block;
border:2px solid #d0d0d0;
border-radius:4px;
-webkit-border-radius:4px;
-moz-border-radius:4px;
box-shadow:inset 0 1px 1px rgba(0, 0, 0, .05);
-webkit-box-shadow:inset 0 1px 1px rgba(0, 0, 0, .05);
-moz-box-shadow:inset 0 1px 1px rgba(0, 0, 0, .05);
}
h2:after{
content:' ';
display:block;
border:1px solid #d0d0d0;
border-radius:4px;
-webkit-border-radius:4px;
-moz-border-radius:4px;
box-shadow:inset 0 1px 1px rgba(0, 0, 0, .05);
-webkit-box-shadow:inset 0 1px 1px rgba(0, 0, 0, .05);
-moz-box-shadow:inset 0 1px 1px rgba(0, 0, 0, .05);
}
pre {
display: block;
padding: 9.5px;
margin: 0 0 10px;
font-size: 13px;
line-height: 20px;
word-break: break-all;
word-wrap: break-word;
white-space: pre;
white-space: pre-wrap;
background-color: #EBEBEB;
border: 1px solid #ccc;
border: 1px solid rgba(0, 0, 0, 0.15);
-webkit-border-radius: 4px;
-moz-border-radius: 4px;
border-radius: 4px;
}
--->
</STYLE>

## Dates and Time in Excel vs. in R

Suppose a scientist were to record in an Excel spreadsheet that a measurement of atmospheric CO2 concentration was taken February 1st, 2016 at 2:00 pm using the flux tower on the roof of CAS. 
One might expect the spreadsheet entry would look something like this:

```{r,echo=FALSE}
knitr::kable(as.data.frame(cbind("Date"="2/1/16", "Time"="2:00:00 PM", "ppm"="403" )))
``` 

One would also assume that the latitude and longitude of BU flux tower would be recorded so that time zone adjustments could be calculated as necessary. 

This is how we are used to looking at dates and times, but unfortunately, this is not the way R formats dates and times. In fact, during the first exercise for this class, you learned that there are R object classes dedicated specifically to date and time.  <br>
To get an idea of how R formats date-time objects, look at how R reports the current time: 
```{r}
Sys.time() #What time is it right now?
```
Here we can see a string with the current date, time and time zone.

What class is this object?
```{r}
class(Sys.time())
```
This returns two classes ("POSIXct" and "POSIXt") which are the two "Date-Time" Classes in R. <br> 
The key to successfully and efficiently using data with date and time components is learning how to convert what you see in your spreadsheets into these date-time formats that R can understand. 

### The goal of this tutorial is to demonstrate how to:
1) Import data that contains date and time components
2) Convert the data into an appropriate R class
3) Make plots with varying date/time axes

The data I will be using for this tutorial is
atmospheric CO2 measurements at Mauna Loa Observatory 
taken directly from the Mauna Loa Observatory [website](http://www.esrl.noaa.gov/gmd/ccgg/trends/full.html) <br>
The data does not come with a time-stamp so I have created a column of randomly generated times for practice. 


## 1) Importing date-time data

First load the data from a csv file:
```{r}
CO2 <-read.csv("Mauna_Loa_Current.csv", stringsAsFactors=FALSE)
```

The first 5 rows of the data set look like this:
```{r, echo=FALSE}
knitr::kable(head(CO2,5))
```

Now we need to learn how to tell R how our data is formatted. <br>
Below are tables of "conversion specifications," i.e. the shorthand one can use to describe date-time formats. This list is by no means exhaustive, and more can be found in the [documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/strptime.html)

### Time 
|Code     |Description                  |Example            |
|--------|----------------------------|------------------|
%H | Hour (00–23). | 14 or 14:00:00 is accepted |
%I | Hour (00–23). | 14 BUT 14:00:00 is NOT accepted |
%p | AM/PM Used with %I and not with %H. | 2 PM |
%M | Minute (00–59).| 01, 40|
%S | Second (00–61) allowing for up to two leap-seconds | 01, 40|

### Dates
|Code     |Description                  |Example            |
|--------|----------------------------|------------------|
|%A       |Full weekday                 |Sunday, Thursday   |
|%a       |Abbreviated weekday          |Sun, Thu           |
|%w       |Weekday (0-6) Sunday = 0 |0, 4 |
|%u       |Weekday (1-7) Sunday = 7 |7, 4  |
|%U       |Week with Sunday as first day (00-53) |21, 27 |
|%W       |Week with Monday as first day (00-53) |21, 27 |  
|%b or %h |Abbreviated month            |May, Jul           |
|%B       |Full month                   |May, July          |
|%d       |Day of the month             |27, 07             |
|%m       |Month  (01-12)                 |05, 12            |
|%j       |Day of the year (001 -366)   |148, 188           |
|%y       |Year without century (00-99) |84, 05  |   
|%Y       |Year with century (00 to 68 prefixed by 20, 69 to 99 prefixed by 19) |1984, 2005 |
|%C       |Century   |19, 20             |

### Common Formats
|Code     |Description                  |Example            |
|--------|----------------------------|------------------|
|%D |Date formatted %m/%d/%y  |05/27/84, 07/07/05 |
|%R | Equivalent to %H:%M | 14:30 |
|%T | Equivalent to %H:%M:%S | 14:30:05 |

### Time zone
Finally, we need to figure out the time zone. <br>
I personally recommend a service called [GeoNames](http://www.geonames.org/)<br>
You can either use the website to determine a time zone by entering latitude and longitude, or you can use the R package `geonames`. To use the R package, you need to have your own personal login, but for this example you can use the name "carya"

```{r,message=FALSE}
require(geonames)
options(geonamesUsername="carya")
gntz <- GNtimezone(lat = CO2$Lat, lng = CO2$Lon, radius = 0)
```

Now we have all the components to convert our date-time data from character strings to date-time objects in R. 

### `strptime()`

Looking back at our own data, we can now see that: 

* Date is `%m/%d/%y`

* Time is `%I:%M:%S %p`

* Time zone is `as.character(gntz$timezoneId)` 

The function `strptime` will convert a character string into a form that R can use for calculations using the conversion specifications.

```{r}
# Paste together the date and time components
CO2$full_date_time <- paste(CO2$Date, CO2$Time) 
# Convert the character string
CO2$full_strp <- strptime(CO2$full_date_time, format = "%m/%d/%y %I:%M:%S %p", tz = as.character(gntz$timezoneId))
```

Now we have date-time entries in a format that R can read!

```{r,echo=F}
knitr::kable(head(data.frame(CO2$Date, CO2$Time, CO2$full_strp),5))
```

Next, we can convert our new data into some of the different Date-Time classes available in R.

## 2) Date-Time Classes

### The Date Class: Dates without time stamps

If your data does not have a time component, then the best class to use is Date. <br>
The Date class is a double precision format (i.e. a numeric value) that represents the number of days since January 1st, 1970. 

```{r}
typeof(as.Date("2/1/2016", format = "%m/%d/%Y"))

#The number of days since January 1st, 1970
cat(as.Date(Sys.time()))
```

`as.Date` is a wrapper for `strptime` so the following commands will return the same thing. 

```{r}
as.Date("1/2/2016", format = "%m/%d/%Y")
as.Date(strptime("1/2/2016", format = "%m/%d/%Y"))
```

`as.Date` can also be used to extract dates from objects with date and time components. 
```{r}
tail(CO2$full_date_time,1)
as.Date(tail(CO2$full_date_time,1), format = "%m/%d/%y %I:%M:%S %p", tz = as.character(gntz$timezoneId))
```

Because a Date object is just a numeric value, it is a popular choice when working with matrices and data frames.

## The POSIXt Classes: Dates with time stamps

### POSIXct -- "ct" for calendar time 

The POSIXct is a double precision format  (i.e. a numeric value) that represents the number of seconds since January 1st, 1970 at 00:00:00 UTC. 

```{r}
typeof(as.POSIXct(Sys.time()))
```


```{r}
# How many seconds has it been since the beginning of 1970
cat(as.POSIXct(Sys.time())) 
```
```{r}
# How many seconds passed between the beginning of 1970 and 1/1/1970 at 00:00:01? 
cat(as.POSIXct("1970-1-1 00:00:01",tz="UTC")) 
```
```{r}
# How many hours has it been since the beginning of 2016?
difftime(as.POSIXct(Sys.time()), as.POSIXct("2016-1-1 00:00:00"), units = "hours")
```

Because POSIXct is just a numeric value, it is a popular choice when working with matrices and data frames. 
```{r}
# Convert our data to POSIXct
CO2$ct <- as.POSIXct(CO2$full_strp)
```

In addition, according to a 2004 editorial by Thomas Lumley in The Newsletter of the R Project, found (here)[https://cran.r-project.org/doc/Rnews/Rnews_2004-1.pdf], 

> The tzone attribute on POSIXlt times
are ignored so it is safer to use POSIXct than
POSIXlt when performing arithmetic or other
manipulations that may depend on time zones.

### POSIXlt -- "lt" for local time (or lists if that helps you remember)

Unlike Date and POSIXct, POSIXlt is a list object that breaks down the date and time into 9 components plus an optional time zone attribute. This not only has the benefit of being human-readable but it also makes it easy to extract specific aspects of a time (such as the day of the week).

```{r}
typeof(as.POSIXlt(Sys.time()))
```


```{r}
# All the entries in the POSIXlt list
knitr::kable(as.data.frame(unclass(as.POSIXlt(Sys.time()))))
```

```{r}
# Extract day of the week (as an integer 1:7, Sunday=7)
as.POSIXlt(Sys.time())$wday
```

```{r}
# Convert our data to POSIXlt
CO2$lt <- as.POSIXlt(CO2$full_strp)
```

## 3) Plotting with dates

Using Date-Time classes makes adjusting axes very simple. You can quickly change the tick marks, as well as the tick mark labels.  
Below are a few examples of different ways to display the Mauna Loa CO2 data. 

```{r}


attach(CO2)

# Simple plot of ppm 
plot(as.Date(full_strp), ppm, type = "l", xlab = "", las=2)
title("Default plot")

# now label every 35 months on the time axis
plot(as.Date(full_strp), ppm, type = "l", xaxt = "n", xlab="")
r <- as.Date(round(range(full_strp), "hours"))
axis.Date(side = 1, at = seq(r[1], r[2], by = "35 months"), format = "%b %Y", las=2)
title("Label every 35 months")

# now label evey 73040 hours (because why not?)
plot(ct, ppm, type = "l", xaxt = "n", xlab = "")
r <- round(range(ct), "hours")
axis.POSIXct(side = 1, at = seq(r[1], r[2], by = "73040 hours"), format = "%D %H:%M", las=2)
title("Label every 73040 hours")

plot(lt$year, ppm)
title("Be careful you don't plot by year instead of label the x axis by year!")


```

