---
title: "Exercise_01"
author: "Tim Condon"
date: "January 31, 2016"
output: html_document
---

1a.
```{r}
log(1)
```
b.
```{r}
log(0)
```
c.
```{r}
log(exp(1))
```
d.
```{r}
log(-5)
```
e.
```{r}
-log(5)
```
f.
```{r}
log(1/5)
```
g. NaNs

2. To convert from radians to degrees.
3. The atan2 function uses two variables while the atan function only uses one.
4. log is the natural logarithm (ln) while log10 computes common logarithms with base 10.
5. 
```{r}
sqrt(5^2+13^2)
```
6. 
```{r}
5*(1+0.04)^pi
```
7.
```{r}
x<-1993/8
x-8
```
8. Sum returns the sum of all vectors present in an argument while cumsum returns a vector with each value the cumulative sum of the elements of the argument. For example, sum of 1:3 gives 8, while cumsum gives the vector 1,3,6.
9.
```{r}
seq(-6,6,by=2)
```
10.
```{r}
seq(-4.8,-3.43,length=8)
```
a.
```{r}
diff(seq(-4.8,-3.43,length=8)
)
```
b.
```{r}
sum(exp(seq(-4.8,-3.43,length=8)
))
```
11. The difference of differences is zero.
```{r}
diff(diff(1:10))
```
12.
```{r}
dat<-read.table('data/Lab1_frogs.txt',header=TRUE)
write.table(dat,'frogs.csv',row.names=FALSE,sep='|')
```
"frogs"|"tadpoles"|"color"|"spots"
1.1|2.03698175474231|"red"|TRUE
1.3|2.87623092770957|"red"|FALSE
```{r}
dat2<-read.table('frogs.csv',header=TRUE,sep='|')
```
13.
```{r}
as.character(dat$spots)
```
14.
```{r}
dat[3:8,1:3]
```
15.
```{r}
dat[1:3,]
```
16.
```{r}
c('Thor','Captain America','Iron Man','Hulk')
```
17.
```{r}
dat[seq(1,nrow(dat),by=2),]
```
18.
a.
```{r}
dat[dat$spots=='TRUE',]
```
b.
```{r}
dat[dat$color=='blue',]
```
c.
```{r}
sum(dat$tadpoles[dat$color=='blue'])
```
d.
```{r}
tads<-subset(dat,dat$tadpoles>3&dat$tadpoles<5)
tads
```
e.
```{r}
red.f<-subset(dat,dat$color=='red')    #first create subset of red frogs
red.f[red.f$frogs<2.5,]
```
f.
```{r}
dat[dat$frogs>5|dat$spots=='FALSE',]
```
19.
a.
```{r}
plot(log(0:3),type='l')
```
b.
```{r}
plot(exp(0:-3),type='l')
```
20.
```{r}
barplot(tapply(dat$frogs,dat$spots,mean))
```
21.
```{r}
blue.f<-subset(dat,dat$color=='blue')
hist(blue.f$frogs)
```
22.
```{r}
apply(dat[,1:2],2,sd)
```
23.
a.
```{r}
red.f<-subset(dat,dat$color=='red')
lmt.f<-lm(red.f$tadpoles~red.f$frogs)
summary(lmt.f)
```
b.
```{r}
plot(red.f$tadpoles~red.f$frogs)
abline(lmt.f)
```
c. The linear model using just frogs is the most complex that is supported by the data.
```{r}
lm.t.f<-lm(dat$tadpoles~dat$frogs)
lm.t.f.s<-lm(dat$tadpoles~dat$frogs+dat$spots)
lm.t.f.s.c<-lm(dat$tadpoles~dat$frogs+dat$spots+dat$color)
summary(lm.t.f)
summary(lm.t.f.s)
summary(lm.t.f.s.c)
```
24.
```{r}
if(all(x>0)) {
  plot(x)  
} else {
  hist(x)
}
```
25.
```{r}
tadpoles<-subset(dat,dat$tadpoles>=5)    #subset for later testing

fancy.plot<-function(x,y,col,z) {
  plot(y~x,cex=1.5,col=as.character(col),pch=z+1,cex.axis=1.3,
       xlab='X Density',ylab='Y Density',cex.lab=1.3)
  abline(lm(y~x),col='green',lwd=3)
}

fancy.plot(tadpoles$frogs,tadpoles$tadpoles,tadpoles$color,tadpoles$spots)
```
26.
```{r}
for( i in 1:10){
  print(i+5)
}
```
27.
```{r}
NT<-20        
N0<-1              
r<-0.2            
N<-rep(N0,NT)
for(t in 1:NT){
    N[t]=N0*(1+r)^t    
}
plot(N)
```
