---
title: "Classic Time-series models"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Grab data

```{r}
library(daymetr)
if(!file.exists("df.Rdata")){
  df <- daymetr::download_daymet(site = "Boston",
                                 lat = 42.36,
                                 lon = -71.06,
                                 start = 2003,
                                 end = 2022,
                                 internal = TRUE)$data
  df$date <- as.Date(paste(df$year,df$yday,sep = "-"),"%Y-%j")
  df$year = as.integer(df$year)
  save(df,file="df.Rdata")
} else {
  load("df.Rdata")
}
```


## Daymet data

```{r}
colnames(df)
train = df |> dplyr::filter(year < 2022) |> dplyr::select(yday,date,tmin..deg.c.)
test  = df |> dplyr::filter(year == 2022) |> dplyr::select(yday,date,tmin..deg.c.)
head(train)
```

## Raw timeseries

```{r}
plot(train$date,train$tmin..deg.c.,type='l')
```

## Null 1: historical means

```{r pressure}
hm = train |> dplyr::group_by(yday) |>
  dplyr::summarise(mean=mean(tmin..deg.c.),sd=sd(tmin..deg.c.))
plot(hm$yday,hm$mean,ylim=c(-15,23),ylab="tmin",xlab="doy",type='l',lwd=3)
lines(hm$yday,hm$mean-1.96*hm$sd,lty=2); lines(hm$yday,hm$mean+1.96*hm$sd,lty=2)
points(test$yday,test$tmin..deg.c.,col=2,pch=18,cex=0.5)
```
## Time series data objects

```{r}
xt <- ts(train$tmin..deg.c.,start=2003,frequency = 365)  ## convert data to a time series object
plot(xt,type='b',ylab="C",xlab="Year",main="Min temperature",cex=0.5)
```


## Null 2: random walk

```{r}
library(forecast)
rw = Arima(xt,c(0,0,0))
rw
```

## Random Walk forecast

```{r}
rwp = forecast(rw,h=365,level=95)
plot(test$date,test$tmin..deg.c.,pch=18,cex=0.5,col=2)
lines(test$date,rwp$mean,lwd=3)
lines(test$date,rwp$lower,lty=2); lines(test$date,rwp$upper,lty=2)
```

## Autocorrelation

```{r}
t    = test$tmin..deg.c.[-nrow(test)]
tlag = test$tmin..deg.c.[-1]
plot(t,tlag,xlab="t",ylab="t+1")
text(-5,15,cor(t,tlag))
```

## Autocorrelation function

```{r}
acf(train$tmin..deg.c.)
```

## Partial Autocorrelation function

```{r}
pacf(train$tmin..deg.c.)
```

## AR(1) timeseries

```{r}
ar1 = Arima(xt,c(1,0,0))
ar1
```
## AR(1) forecast

```{r}
ar1p = forecast(ar1,h=365,level=95)
plot(test$date,test$tmin..deg.c.,pch=18,cex=0.5,col=2)
lines(test$date,ar1p$mean,lwd=3)
lines(test$date,ar1p$lower,lty=2); lines(test$date,ar1p$upper,lty=2)
```

## Auto.arima: AR only

```{r}
ar = auto.arima(xt,max.p=10,max.d = 0,max.q = 0,seasonal=FALSE)
ar
```

## AR forecast

```{r}
arp = forecast(ar,h=365,level=95)
plot(test$date,test$tmin..deg.c.,pch=18,cex=0.5,col=2)
lines(test$date,arp$mean,lwd=3)
lines(test$date,arp$lower,lty=2); lines(test$date,arp$upper,lty=2)
```

## Detrending

```{r}
rt <- ts(train$tmin..deg.c. - hm$mean[train$yday],start=2003,frequency = 365)  ## convert data to a time series object
plot(rt,ylab="C",xlab="Year",main="Min temperature anomaly",cex=0.5)
```

## Autocorrelation

```{r}
par(mfrow=c(1,2))
acf(rt)
pacf(rt)
```

## auto.arima

```{r}
rr = auto.arima(rt,max.p=10,max.d = 0,max.q = 0,seasonal=FALSE)
rr
```

## Forecast of anomalies

```{r}
rrp = forecast(rr,h=365,level=95)
test.anomaly = test$tmin..deg.c. - hm$mean
plot(test$date,test.anomaly,pch=18,cex=0.5,col=2)
lines(test$date,rrp$mean,lwd=3)
lines(test$date,rrp$lower,lty=2); lines(test$date,rrp$upper,lty=2)
```

## actual forecast

```{r}
plot(test$date,test$tmin..deg.c.,pch=18,cex=0.5,col=2)
lines(test$date,rrp$mean+hm$mean,lwd=3)
lines(test$date,rrp$lower+hm$mean,lty=2); lines(test$date,rrp$upper+hm$mean,lty=2)
```

## predictive uncertainties

```{r}
plot(test$date,rwp$upper-rwp$lower,type='l',
     ylim=c(0,40),xlim=as.Date(c("2022-01-01","2022-03-01")),
     xlab="Time",ylab="CI width")
lines(test$date,hm$sd*1.96*2,col=2)
lines(test$date,ar1p$upper-ar1p$lower,col=3)
lines(test$date,arp$upper-arp$lower,col=4)
lines(test$date,rrp$upper-rrp$lower,col=5)
legend("bottomright",col=1:5,lwd=2,legend=c("RandWalk","HistMeans","AR(1)","AR(auto)","detrended AR(auto)"))
```

